Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-06-20 paper.tex
XCR update

\C{We compare the calculated fluxes to 95\% c.l. gamma-ray upper
  limits set by Fermi-LAT after 18 months of observations
  \cite{2010ApJ...717L..71A}. In particular, to achieve a more
  reliable comparison we adopt the maximal spatially extended limits
  (if nothing else is stated) since the gamma-ray flux from our
  brightest clusters all have an angular extent above 1 degree on the
  sky when the boost from substructures is included. In fact, current
  Fermi flux upper limits are not derived for the kind of extended
  emission that we get with our treatment of substructures, hence the
  Fermi limits may not be adequate for some of these clusters. This
  and other recent work highlights that the improved substructure
  models imply that Fermi limits will have to be re-calculated to
  accommodate for the large source extensions. Note, however, that for
  most clusters the assumption of a point source is well justified.}

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Fornax.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Coma.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Virgo.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Perseus.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\caption{\it Comparing the gamma-ray emission from different
  clusters. We show the differential flux of clusters \C{in the upper
    panel of each figure}: Fornax (upper left), Coma (upper right),
  Virgo (lower left), and Perseus (lower right). We show CR induced
  emission (red solid), a leptophilic (LP) model that includes both
  final state radiation and IC upscattered CMB, dust and starlight
  (green dash-dotted), and the benchmark $\Kp$ model (BM) that
  includes continuum emission, and IC upscattered CMB, dust and
  starlight (blue dashed). The arrows show the spatially extended
  differential upper limits set by Fermi-LAT in the energy ranges
  $0.2-1$~GeV, $1-10$~GeV, and $10-100$~GeV from left to right. We
  show the individual components for the BM model and LP model for
  Virgo and Perseus, respectively. \C{In the lower panel of each
    figure we show differential spectral index where $\dd F_\gamma/\dd
    E_\gamma \sim E_\gamma^{-\alpha_\gamma}$.} The flux from the
  clusters is integrated out to $\rvir$ using a point spread function
  of $0.1\deg$ and includes the boost from substructures. We find that
  the lower GeV energy regime is most constraining due to the peak
  sensitivity of Fermi at these energies and use this regime to get
  upper limits on boost factors.}
 \label{fig:clu_comp}
\end{minipage}
\end{figure*}

@@

\C{A quantity that is of great scientific interest is the CR pressure
  ($P_\CR$) relative to the thermal pressure ($P_\rmn{th}$) as it
  directly bias galaxy cluster observables and the resulting
  cosmological constraints, such as the hydrostatic mass bias or the
  Sunyaev-Zel’dovich signal measured by Planck, South Pole Telescope,
  or Atacama Cosmology Telescope. In the following we estimate the
  mass scaling of the relative CR pressure, $X_\CR =
  P_\CR/P_\rmn{th}$, using the CR formalism outlined in
  \cite{2007A&A...473...41E,2010MNRAS.409..449P}.  First we calculate
  the CR distribution function $f = C\,p^{-\alpha}$, where $C$ is the
  normalization of each power-law, $p=P/m_\rmn{p} c$ is the normalized
  CR proton momentum, and $\alpha$ is the spectral index. The CR
  pressure $P_\CR \propto C=\tilde{C}/\rho_\rmn{gas}$, where
  $\tilde{C}\propto M^{0.44}$ for clusters with a mass $\mvir\gtrsim
  10^{14}\,\msun$ \cite{2010MNRAS.409..449P}. We determine the
  mass scaling of the gas density through its relation to the enclosed
  gas fraction $f_\rmn{gas}$, and find that $\rho_\rmn{gas}\propto
  f_\rmn{gas} \propto M^{0.135}$ \cite{2009ApJ...693.1142S}. The
  thermal pressure is derived from the temperature in virial
  equilibrium, and scales as $P_\rmn{th}\propto k_\B T \propto G M/R
  \propto M^{0.67}$. Finally we can derive the mass scaling of the
  volume weighted relative CR pressure from
\begin{equation}
 <X_\CR> = \frac{<P_\CR>}{<P_\rmn{th}>} \propto \frac{<\rho_\rmn{gas}\,
   \tilde{C}>}{<k_\B T>} \propto M^{-0.06}\,,
\end{equation}
which is roughly a constant, i.e. almost independent on mass. Hence we
adopt a constant $<X_\CR>\approx 0.02$ derived from a representative
simulated cluster \cite{2008MNRAS.385.1211P}. Since the CRs that build
up the bulk of the CR pressure also give rise to gamma-ray emission
through the production of decaying pions, we use gamma-ray upper
limits from Fermi-LAT to constrain the volume weighted $X_\CR$. In
Fig.~\ref{fig:XCR} we show the volume weighted relative CR pressure
for 32 clusters, for which LAT upper limits that assume King profiles
are derived. For most of the clusters in the sample we can constrain
$X_\CR < (0.1-0.2)$ using the extended gamma-ray upper limits, where
the most constraining clusters are Norma and Coma with $X_\CR <
0.07$. Most of the 32 clusters have $X_\CR$ upper limits within a
factor few from the best constraining clusters, which suggests that
stacking of clusters will be able to improve these limits. This might
even be improved further using a scaled stacking with our model as a
prior. In previous work \cite{2010ApJ...717L..71A} where the 1.5 year
Fermi data was used to derive limits on $X_\CR$ they found on average
more constraining limits on the most massive clusters, while we find
better limits on the low mass clusters. This difference is explained
by the different modelling of $P_\CR$; they assumed a mass
independent CR pressure while we assume that $P_\CR\sim
M^{0.44}$. Furthermore, the gamma-ray flux predictions that we use in
this work to constrain $X_\CR$ assumes the true gas density
distribution for each cluster \footnote{Derived from the parameters in
  the extended HIFLUGCS catalogue.} that accounts for
cluster-to-cluster scatter, in contrast to the mass-to-luminosity
scaling relations that was used in previous works.}

@@

We substantially improve the modelling of the expected gamma-ray
signal from pion-decay resulting from hadronic CR interactions with
gas protons over previous work \citep{2010ApJ...717L..71A}. We employ
the analytic model of the CR spectrum and spatial distribution,
recently developed by \citet{2010MNRAS.409..449P} that is based on
cosmological hydrodynamical simulations of galaxy clusters that
self-consistently follow the evolution of CRs in cosmic structure
formation. Adopting the model for all clusters in the HIFLUGCS sample
of flux-limited X-ray clusters, we compare the expected gamma-ray
emission to that inferred from using $L_\gamma$-cluster mass scaling
relation. We note that the predictions of our ``optimistic'' model are
perfectly in agreement with Fermi upper limits for all clusters.  The
hint of a tension that the scaling relation modeling showed for Norma
\citep{2010ApJ...717L..71A} could not be confirmed due to the lower
gas density of this cluster relative to the average.  As expected, the
analytical CR model accounts for the ``scatter'' to the scaling
relation and biases the gamma-ray flux high for prominent cool core
clusters by up to a factor of 6 relative to the expectation from the
scaling relation. This is due to the high target gas densities and
associated CR densities in those cool cores that shape the spatial
emission characteristic to be very similar to that observed in thermal
X-rays. We caution that these predictions only apply for the case of
negligible active CR transport such as CR streaming and diffusion
relative to the gas. Gamma-ray fluxes might be considerably smaller in
those cluster that do not show an extended diffuse radio(-mini) halo
that might point to a centrally concentrated CR distribution
\citep{2011A&A...527A..99E}. The CR spectrum $E^2 dN/dE$ shows the
characteristic pion bump at energies around 1~GeV. It is expected to
dominate the DM annihilation signal of BM models for all
clusters/groups but Fornax. \C{We conclude that due to the large
  spatial extension of the brightest clusters it is not possible for
  Fermi to rule out the LP model nor to constrain the BM DM models
  without improved modelling. However, we can put tight constraints on
  the boost from substructures and Sommerfeld enhancement using the LP
  model. Hence, clusters are a nice complement to direct and
  accelerator searches for the nature of DM.}\del{We conclude that it
  will be possible to rule out the LP model with six years of Fermi
  data and start to constrain some BM DM models. Hence this should
  nicely complement direct and accelerator searches for the nature of
  DM.}

>>>>

\C{We compare the calculated fluxes to gamma-ray upper limits (95\%
  c.l.) set by Fermi-LAT after 18 months of observations
  \cite{2010ApJ...717L..71A}. In particular, to achieve a more
  reliable comparison we adopt, if nothing else is stated, the maximal
  spatially extended limits since the gamma-ray flux from our
  brightest clusters all have an angular extent above 1 degree on the
  sky when the boost from substructures is included. In fact, current
  Fermi flux upper limits are not derived for the kind of extended
  emission that we get with our treatment of substructures, hence the
  Fermi limits may not be adequate for some of these clusters. This
  and other recent work highlights that the improved substructure
  models imply that Fermi limits will have to be re-calculated to
  accommodate for the large source extensions. Note, however, that for
  most clusters the assumption of a point source is well justified.}

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Fornax.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Coma.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Virgo.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Perseus.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\caption{\it Comparing the gamma-ray emission from different
  clusters. We show the differential flux of clusters \C{in the upper
    panel of each figure}: Fornax (upper left), Coma (upper right),
  Virgo (lower left), and Perseus (lower right). We show CR induced
  emission (red solid), a leptophilic (LP) model that includes both
  final state radiation and IC upscattered CMB, dust and starlight
  (green dash-dotted), and the benchmark $\Kp$ model (BM) that
  includes continuum emission, and IC upscattered CMB, dust and
  starlight (blue dashed). The arrows show the spatially extended
  differential upper limits set by Fermi-LAT in the energy ranges
  $0.2-1$~GeV, $1-10$~GeV, and $10-100$~GeV from left to right. We
  show the individual components for the BM model and LP model for
  Virgo and Perseus, respectively. \C{In the lower panel of each
    figure we show differential spectral indices, $\alpha_\gamma$,
    where $\dd F_\gamma/\dd E_\gamma \sim E_\gamma^{-\alpha_\gamma}$.}
  The flux from the clusters is integrated out to $\rvir$ using a
  point spread function of $0.1\deg$ and includes the boost from
  substructures. We find that the lower GeV energy regime is most
  constraining due to the peak sensitivity of Fermi at these energies
  and use this regime to get upper limits on boost factors.}
 \label{fig:clu_comp}
\end{minipage}
\end{figure*}

@@

\C{A quantity that is of great scientific interest is the CR pressure
  ($P_\CR$) relative to the thermal pressure ($P_\rmn{th}$) as it
  directly bias galaxy cluster observables and the resulting
  cosmological constraints, such as the hydrostatic mass bias or the
  Sunyaev-Zel’dovich signal measured by Planck, South Pole Telescope,
  or Atacama Cosmology Telescope. In the following we estimate the
  mass scaling of the relative CR pressure, $X_\CR =
  P_\CR/P_\rmn{th}$, using the CR formalism outlined in
  \cite{2007A&A...473...41E,2010MNRAS.409..449P}. We derive the CR
  pressure from the CR distribution function $f = C\,p^{-\alpha}$,
  where $C$ is the normalization of each power-law, $p=P/m_\rmn{p} c$
  is the normalized CR proton momentum, and $\alpha$ is the CR
  spectral index. The pressure is given by $P_\CR \propto
  C=\tilde{C} \,\rho_\rmn{gas}$, where $\tilde{C}\propto M^{0.44}$ for
  clusters with a mass $\mvir\gtrsim 10^{14}\,\msun$
  \cite{2010MNRAS.409..449P}. We determine the mass scaling of the gas
  density through its relation to the enclosed gas fraction
  $f_\rmn{gas}$, and find that $\rho_\rmn{gas}\propto f_\rmn{gas}
  \propto M^{0.135}$ \cite{2009ApJ...693.1142S}. The thermal pressure
  is derived from the temperature in virial equilibrium, and scales as
  $P_\rmn{th}\propto k_\B T \propto G M/R \propto M^{0.67}$. Finally
  we can derive the mass scaling of the volume weighted relative CR
  pressure from
\begin{equation}
 <X_\CR> = \frac{<P_\CR>}{<P_\rmn{th}>} \propto \frac{<\rho_\rmn{gas}\,
   \tilde{C}>}{<k_\B T>} \propto M^{-0.1}\,,
\end{equation}
which is roughly a constant, i.e. almost independent on mass. Hence we
adopt a constant $<X_\CR>\approx 0.02$ derived from a representative
simulated cluster \cite{2008MNRAS.385.1211P}. Since the CRs that build
up the bulk of the CR pressure also give rise to gamma-ray emission
through the production of decaying pions, we use gamma-ray upper
limits from Fermi-LAT to constrain the volume weighted $X_\CR$. In
Fig.~\ref{fig:XCR} we show the volume weighted relative CR pressure
for 32 clusters, for which LAT upper limits that assume King profiles
are derived. For most of the clusters in the sample we can constrain
$X_\CR < (0.1-0.2)$ using the extended gamma-ray upper limits, where
the most constraining clusters are Norma and Coma with $X_\CR <
0.07$. Most of the 32 clusters have $X_\CR$ upper limits within a
factor few from the best constraining clusters, which suggests that
stacking of clusters will be able to improve these limits. This might
even be improved further using a scaled stacking with our model as a
prior. In the recent work \cite{2010ApJ...717L..71A} where the 1.5
year Fermi data was used to derive limits on $X_\CR$, they found that
the most massive clusters are on average more constraining, while we
find better limits on the low mass clusters. This difference is
explained by the different priors of $P_\CR$; they assumed a mass
independent CR pressure derived from a single CR power-law
distribution function, while we adopt a prior infered from cosmological
simulations that dynamically trace the CRs during cluster assembly and
allow for a varyning spectral index where $P_\CR\sim
M^{0.44}$. Furthermore, the gamma-ray flux predictions that we use in
this work to constrain $X_\CR$ assumes the true gas density
distribution for each cluster \footnote{Derived from the parameters in
  the extended HIFLUGCS catalogue.} that accounts for
cluster-to-cluster scatter, in contrast to the mass-to-luminosity
scaling relations that was used in previous work.}

@@

We substantially improve the modelling of the expected gamma-ray
signal from pion-decay resulting from hadronic CR interactions with
gas protons over previous work \citep{2010ApJ...717L..71A}. We employ
the analytic model of the CR spectrum and spatial distribution,
recently developed by \citet{2010MNRAS.409..449P} that is based on
cosmological hydrodynamical simulations of galaxy clusters that
self-consistently follow the evolution of CRs in cosmic structure
formation. Adopting the model for all clusters in the HIFLUGCS sample
of flux-limited X-ray clusters, we compare the expected gamma-ray
emission to that inferred from using $L_\gamma$-cluster mass scaling
relation. We note that the predictions of our ``optimistic'' model are
perfectly in agreement with Fermi upper limits for all clusters.  The
hint of a tension that the scaling relation modeling showed for Norma
\citep{2010ApJ...717L..71A} could not be confirmed due to the lower
gas density of this cluster relative to the average.  As expected, the
analytical CR model accounts for the ``scatter'' to the scaling
relation and biases the gamma-ray flux high for prominent cool core
clusters by up to a factor of 6 relative to the expectation from the
scaling relation. This is due to the high target gas densities and
associated CR densities in those cool cores that shape the spatial
emission characteristic to be very similar to that observed in thermal
X-rays. We caution that these predictions only apply for the case of
negligible active CR transport such as CR streaming and diffusion
relative to the gas. Gamma-ray fluxes might be considerably smaller in
those cluster that do not show an extended diffuse radio(-mini) halo
that might point to a centrally concentrated CR distribution
\citep{2011A&A...527A..99E}. The CR spectrum $E^2 dN/dE$ shows the
characteristic pion bump at energies around 1~GeV. It is expected to
dominate the DM annihilation signal of BM models for all
clusters/groups but Fornax. \C{We conclude that due to the large
  spatial extension of the brightest clusters it is not possible for
  Fermi to rule out the LP model or to constrain the BM DM models
  without improved modelling. However, we can put tight constraints on
  the boost from substructures and Sommerfeld enhancement using the LP
  model. Hence, clusters are a nice complement to direct and
  accelerator searches for the nature of DM.}\del{We conclude that it
  will be possible to rule out the LP model with six years of Fermi
  data and start to constrain some BM DM models. Hence this should
  nicely complement direct and accelerator searches for the nature of
  DM.}
